EagerNet: Early Predictions of Neural Networks for Computationally Efficient Intrusion Detection

### 2020-07-26 paper/paper.tex
improvements based on feedback

\section{Introduction}
% what are fcnn
\glspl{fcnn} have gained remarkable attention in recent years due to the availability of data and computational resources for research. Many state-of-the-art architectures currently used in different applications are effectively based on simple \gls{fcnn} architectures. Experiments show that the more layers (and neurons) the network includes, the more desirable it results in better learning performance \cite{eldan_power_2016}.

@@

%EagerNet
To solve this, we propose a new architecture that, on the one hand, uses the entire network capacity for learning network intrusions and, on the other hand, stops the forward-pass and makes predictions as soon as confidence reaches a certain threshold. In other words, it makes it possible to avoid evaluating the entire network with all layers for some samples and to stop the neural network evaluation at a particular layer if the achieved confidence at that point is high enough. We call different \glspl{fcnn} architectures that uses the same proposed approach \emph{EagerNets} (Eager Stopping Networks). EagerNets therefore allow, where possible, the reduction of computing resources and energy usage while achieving detection performance comparable to that of a full forward-pass.

% conclusion
%\todo{Max: For such a short paper it might not be necessary to give an overview?}
%\todo{Fares: I now added just a sentence}
In the following sections, we present the existing related work, introduce the concept behind EagerNet and finally show and analyze the evaluation results based on two network intrusion detection datasets.

 \begin{figure}
  \includegraphics[width=\linewidth]{figures/fcnn_new.pdf}
  \caption{Conventional \gls{fcnn} architecture.}
  \label{fig:fcnn}
\end{figure}

@@

The concept of \textit{cascading} classifiers encompasses several classifiers that are used in a cascade (like a chain) \cite{viola_robust_2004, xu_classifier_2014, wang_efficient_2015}. If accuracy of a classifier in the beginning is good enough, then there is no need to evaluate the other classifiers in the chain (cascade), which saves computational resources. This approach was predominantly explored for classical machine learning methods. Two possibilities are to either place all classifiers in a row or in the shape of tree. If a tree is used, different classifiers can be used at each step, depending on which features are probably relevant for the sample currently under investigation. The difference to our approach is that we exclusively consider neural networks as the classifier and that our approach is significantly more straightforward to implement.

@@

The major difference between all existing works and ours is simplicity. While many of the previously proposed works report good results, they usually require advanced math knowledge to be implemented, which hinders their deployment in real-world scenarios. These also often contain components such as additional output layers, which add computational overheads that are shown not to be necessary.

@@

\gls{ml} has been widely used in the last decade for network traffic analysis and specifically anomaly detection systems. Many of the works proposed \gls{ids} architectures based on well-known supervised techniques \cite{buczak_survey_2016,berman_survey_2019}. The concept of having a classifier trained on pre-stored attack patterns and using it to predict similar behaviours is commonly used whereby remarkable success has so far been achieved \cite{vigneswaran_evaluating_2018,ferrag_deep_2020,almseidin_evaluation_2017}. In particular, \glspl{fcnn} are commonly utilized whenever large amounts of network traffic are available. In order to present network data to such architectures, a number of traffic representations have been proposed depending on the application. The statistical representation of flow characteristics is so far the most widely used. In this work, we define a network flow to be the exchange of packets between two end-hosts. Packets can be identified and aggregated using the five-tuple key: \emph{sourceIP}, \emph{destinationIP}, \emph{sourceTransportPort}, \emph{destinationTransportPort} and \emph{protocolIdentifier}. Thereafter, packet header features are extracted and statistical combinations are computed. \autoref{tab:features} shows the CAIA \cite{williams_preliminary_2006} network traffic representation. It consists of 12 features, 7 of which are measured in both directions and 2 of which are expanded using four statistical combinations (i.e. \emph{A} becomes \emph{mean(A)}, \emph{min(A)}, \emph{max(A)} and \emph{stdev(A)}). We use CAIA for all experiments in this paper and set an observation timeout of 1,800 seconds, after which we terminate flows. Each flow is considered to be one data sample and has two sets of labels: a binary (attack/benign) label and also the attack family (one binary label for each attack family using one-hot-encoding). The ultimate goal is that after a training process, the trained \gls{ids} can correctly classify new flows and also determine what kind of attack family they belong to.

@@

Because some network attack families are simpler to detect compared to others we modify the standard \gls{fcnn} architecture so that for easily classifiable samples the network doesn't have to evaluate all layers. Our novel architecture is based on the assumption that deep neural networks with more layers can learn increasingly complex functions which in turn are only required for the classification of some particular notoriously hard-to-classify samples. As a result, we build a network such that an additional set of neurons is connected to each layer (a copy of the output layer), allowing for direct predictions at each layer. The EagerNet architecture is shown in \autoref{fig:eagerNet_forward}. Initial neurons are shown in gray and output neurons (per layer) are shown in green. Once a network flow is observed and features are extracted, a data sample $\textbf{x}$ is fed to the network via the input layer. The neural network proceeds evaluating the layers one by one and yields a prediction confidence value at each layer. We define the confidence as the value that represents how sure the network is about a certain decision or sample belonging to a certain class. It is a number between 0.5 and 1 that is obtained by applying the sigmoid function \cite{noauthor_sigmoid_2020} on the output neuron and computing how far the result is from the farthest label (e.g. if the prediction scalar of a sample is 0.2, the farthest label is 1 and therefore a confidence of 0.8 that the sample belongs to class 0). By looking at the confidence the neural network determines whether additional layers should be processed or if the currently obtained confidence is high enough. This methodology ensures that simpler samples are classified early in the forward-pass and that more complicated samples are passed into deeper layers. In the section \ref{subsubsec:confidence-speed-tradeoff}, we show that EagerNet reduces resource consumption on average, while achieving comparable performance compared to normal \gls{fcnn}.

@@

For the multiclass approach, on the other side, there is one output neuron per attack family, and each neuron's output indicates how likely it is, that the currently processed sample belongs to the given attack family. During back-propagation, we use the categorical cross-entropy loss to penalize all outputs together, so there is can be no overlap between classes. \autoref{eq:cceloss} shows how the loss is determined for a batch of size $N$, with the $n^{th}$ input $x_{n}$ and the softmax activation function \cite{noauthor_softmax_2020} $\phi(.)$, $C$ being the number of classes or output neurons and $i$ is the current class/output.

@@

\section{Evaluation and Discussion}
\label{evaluation_and_discussion}
In this section we discuss the different experimental results and observations. Overall, EagerNet shows promising performance that are comparable to \glspl{fcnn} with the ability to save resources. The interested reader can reproduce experiments with the provided sources in our repository\footnote{\url{github.com/CN-TU/ids-backdoor/tree/eager}}.

@@

\subsubsection{Weighted loss}
In \autoref{tab:weights} we gather results from different networks trained using different numbers of hidden layers and weight distributions each with 128 neurons per layer. In addition, we compare the results using four metrics (Accuracy, Precision, Recall and Youden's J) and show the performance of the last layer (by setting the confidence level to the maximum). The CICIDS2017 indices %\todo{Max: Indices? Fares: it's commun to refer to acc, prec, recall etc. as indices/scores}
show that the weight distribution and the architecture itself (number of layers) have little effect on performance. In fact, since the last layer is always trained with one loss regardless of the weight distribution, only the first few layers are affected by the combined-loss learning procedure. Nonetheless, since the accuracy did not improve when increasing the number of layers, that implies that most network patterns are already separable after the few first layers which makes the classification task for CICIDS2017 manageable even for shallow neural networks. Results of UNSW-NB15 show a slightly different behavior. %\todo{Max: Sure? All the results look the same? Fares: slightly better}
The uniform weighting of losses gives best scores on average and, in addition, deeper architectures outperform the shallow ones. This implies that the network patterns in this specific dataset are complex than in the previous one therefore, requiring more non-linearities (layers) to find the optimal \emph{input:prediction} mapping function.

@@

\autoref{fig:confidence} shows the accuracy-speed balance in terms of confidence threshold. Four different networks for both datasets are shown. We experiments with two networks: 128 neurons by 3 layers and 64 neurons by 12 layers. The confidence threshold is shown on the horizontal axis, the average number of layers used over all samples and the average accuracy achieved are both shown on the vertical axis. Curves show promising and consistent results, whereby the network always uses few layers for the majority of samples and only continues to evaluate further layers if the confidence level required is too high or the sample is noisy thus it is difficult to make correct decisions in early layers. We observe moreover two phenomena:
\begin{itemize}
\item The accuracy of CICIDS2017 increases almost linearly with respect to confidence and reaches its maximum when all layers are evaluated.
\item The accuracy of UNSW-NB15 is stable when the confidence level is increased until $\approx 0.98$ where it leaps noticeably once a further layer is evaluated. This suggests that some patterns were ``learned'' only on that specific layer and hence, respective samples suddenly were correctly classified.
\end{itemize}
Overall, EagerNet confirms that it is possible to trade a tiny percentage of accuracy in order to save a significant amount of resources. In section \ref{subsubsec:comparability_to_fnns}, we see that, it is feasible to achieve comparable accuracies to traditional \gls{fcnn} allowing the reduction of costs.

@@

In \autoref{tab:comparability}, we show the F1 score obtained at the last layer of two architectures: Simple \gls{fcnn} and EagerNet with the same number of layers and neurons. We use the F1 score to reduce the class imbalance bias. Results suggest that on average, EagerNet achieves similar scores to a conventional \gls{fcnn} at the last layer across all variants and datasets. It is worth mentioning, however, that EagerNet is used in this case as a conventional neural network and predictions are taken only from the last layer. Using predictions from various layers by setting a confidence threshold might lead to higher accuracy if middle layers are optimized to predict certain classes.

@@

\begin{figure*}
\center
\begin{tabular}{cccc}
\subfloat[CICIDS2017 (12 layers $\times$ 64 neurons). \todo{Fares: Model is still running to generate a better figure.}]{\includegraphics[width=0.55\columnwidth]{figures/multiclass17.pdf}} &
\subfloat[CICIDS2017 (5 layers $\times$ 128 neurons).]{\includegraphics[width=0.35\columnwidth]{figures/multiclass17-3.pdf}} 
\subfloat[UNSW-NB15 (12 layers $\times$ 64 neurons).]{\includegraphics[width=0.68\columnwidth]{figures/multiclass15.pdf}} &
\subfloat[UNSW-NB15 (5 layers $\times$ 128 neurons).]{\includegraphics[width=0.38\columnwidth]{figures/multiclass15-3.pdf}}
\end{tabular}
\caption{Accuracy achieved on a test set per layer and attack family (sorted vertically). Some attacks have poor accuracy because only few samples are observed (refer to \autoref{fig:datasets}). Forcing the network to learn those samples might lead to over-fitting.}
\label{fig:multiclass}
\end{figure*}

>>>>

\section{Introduction}
% what are fcnn
\glspl{fcnn} have gained remarkable attention in various domains in recent years due to the availability of data and computational resources for research. Many state-of-the-art architectures currently used in different applications are effectively based on simple \gls{fcnn} architectures. Since the size of a neural network cannot be mathematically defined, experts use rule-of-thumb to define the best architecture that fits their needs. Yet, experiments show that the more layers (and neurons) the network includes, the better its learning performance becomes \cite{eldan_power_2016}.

@@

%EagerNet
To obtain early predictions, we propose a new architecture that, on the one hand, uses the entire network capacity for learning network intrusions and, on the other hand, stops the forward-pass and makes predictions as soon as confidence reaches a certain threshold. In other words, it makes it possible to avoid evaluating the entire network with all layers for some samples and to stop the neural network evaluation at a particular layer if the achieved confidence at that point is high enough. We call different \glspl{fcnn} architectures that uses the same proposed approach \emph{EagerNets} (Eager Stopping Networks). EagerNets therefore allow, where possible, the reduction of computing resources and energy usage while achieving detection performance comparable to that of a full forward-pass.

 \begin{figure}
  \includegraphics[width=\linewidth]{figures/fcnn_new1.pdf}
  \caption{Conventional \gls{fcnn} architecture.}
  \label{fig:fcnn}
\end{figure}
% conclusion
%\todo{Max: For such a short paper it might not be necessary to give an overview?}
%\todo{Fares: I now added just a sentence}
In the following sections, we present the existing related work, introduce the concept behind EagerNet and finally show and analyze the evaluation results based on two network intrusion detection datasets.

@@

The concept of \textit{cascading} classifiers encompasses several classifiers that are used in a cascade (like a chain) \cite{viola_robust_2004, xu_classifier_2014, wang_efficient_2015}. If accuracy of a classifier in the beginning is good enough, then there is no need to evaluate the other classifiers in the chain (cascade), which saves computational resources. This approach was predominantly explored for classical machine learning methods. Two possibilities are to either place all classifiers in a row or in the shape of a tree. If a tree is used, different classifiers can be used at each step, depending on which features are probably relevant for the sample currently under investigation. The difference to our approach is that we exclusively consider neural networks as the classifier and that our approach is significantly more straightforward to implement.

@@

The major difference between all existing works and ours is simplicity in terms of implementation complexity and interpretability. While many of the previously proposed works report good results, they usually require advanced math knowledge to be implemented, which hinders their deployment in real-world scenarios. These also often contain components such as additional output layers, which add computational overheads that are shown not to be necessary.

To the best of our knowledge, our work considers for the first time such an approach for network stream data to obtain early flow classification yet with remarkable performance.

@@

\gls{ml} has been widely used in the last decade for network traffic analysis and specifically anomaly detection systems. Many of the works proposed \gls{ids} architectures based on well-known supervised techniques \cite{buczak_survey_2016,berman_survey_2019}. The concept of having a classifier trained on pre-stored attack patterns and using it to predict similar behaviours is commonly used whereby remarkable success has so far been achieved \cite{vigneswaran_evaluating_2018,ferrag_deep_2020,almseidin_evaluation_2017}. In particular, \glspl{fcnn} are commonly utilized whenever large amounts of network traffic are available. In order to present network data to such architectures, a number of traffic representations have been proposed depending on the application. The statistical representation of flow characteristics is so far the most widely used. In this work, we define a network flow to be the exchange of packets between two end-hosts. Packets can be identified and aggregated using the five-tuple key: \emph{sourceIP}, \emph{destinationIP}, \emph{sourceTransportPort}, \emph{destinationTransportPort} and \emph{protocolIdentifier}. Thereafter, packet header features are extracted and statistical combinations are computed. \autoref{tab:features} shows the CAIA \cite{williams_preliminary_2006} network traffic representation. It consists of 12 features, 7 of which are measured in both directions and 2 of which are expanded using four statistical combinations (i.e. \emph{A} becomes \emph{mean(A)}, \emph{min(A)}, \emph{max(A)} and \emph{stdev(A)}). We use CAIA for all experiments in this paper and set an observation timeout of 1,800 seconds (slightly longer than the longest attack activity in the dataset), after which we terminate flows. Each flow is considered to be one data sample and has two sets of labels: a binary (attack/benign) label and also the attack family (one binary label for each attack family using one-hot-encoding). The ultimate goal is that after a training process, the trained \gls{ids} can correctly classify new flows and also determine what kind of attack family they belong to.

@@

Because some network attack families are simpler to detect compared to others we modify the standard \gls{fcnn} architecture so that for easily classifiable samples the network doesn't have to evaluate all layers. Our novel architecture is based on the assumption that deep neural networks with more layers can learn increasingly complex functions which in turn are only required for the classification of some particular notoriously hard-to-classify samples. As a result, we build a network such that an additional set of neurons are connected to each layer (a copy of the output neurons instead of an entire meta-network), allowing for direct predictions at each layer. The EagerNet architecture is shown in \autoref{fig:eagerNet_forward}. Initial neurons are shown in gray and output neurons (per layer) are shown in green. Once a network flow is observed and features are extracted, a data sample $\textbf{x}$ is fed to the network via the input layer. The neural network proceeds evaluating the layers one by one and yields a prediction confidence value at each layer. We define the confidence as the value that represents how sure the network is about a certain decision or sample belonging to a certain class. It is a number between 0.5 and 1 that is obtained by applying the sigmoid function \cite{noauthor_sigmoid_2020} on the output neuron and computing how far the result is from the farthest label (e.g. if the prediction scalar of a sample is 0.2, the farthest label is 1 and therefore a confidence of 0.8 that the sample belongs to class 0). By looking at the confidence the neural network determines whether additional layers should be processed or if the currently obtained confidence is high enough. This methodology ensures that simpler samples are classified early in the forward-pass and that more complicated samples are passed into deeper layers. In the section \ref{subsubsec:confidence-speed-tradeoff}, we show that EagerNet reduces resource consumption on average, while achieving comparable performance compared to normal \gls{fcnn}.

\subsubsection{Network Architecture}
For our experiments, we use a neural network that consists of as many layers as needed with Leaky ReLU activations \cite{maas_rectifier_2013} ($\alpha = 0.1$) after each layer. Moreover and to reduce over-fitting, we use neurons dropout ($r=0.2$).

@@

For the multiclass approach, on the other side, there is one output neuron per attack family, and each neuron's output indicates how likely it is, that the currently processed sample belongs to the given attack family. During back-propagation, we use the categorical cross-entropy loss to penalize all outputs together, so there can be no overlap between classes. \autoref{eq:cceloss} shows how the loss is determined for a batch of size $N$, with the $n^{th}$ input $x_{n}$ and the softmax activation function \cite{noauthor_softmax_2020} $\phi(.)$, $C$ being the number of classes or output neurons and $i$ is the current class/output.

@@

\section{Evaluation and Discussion}
\label{evaluation_and_discussion}
In this section we discuss the different experimental results and observations. Overall, EagerNet shows promising performance that are comparable to \glspl{fcnn} that always use all layers yet with the ability to save resources. The interested reader can reproduce experiments with the provided sources in our repository\footnote{\url{github.com/CN-TU/ids-backdoor/tree/eager}}.

@@

In what follows, we show the performance of the EagerNet architecture taking into account various aspects. We stop training all models after 800 epochs to ensure fair comparison and we monitor both training and validation losses to prevent over-fitting.

In \autoref{tab:comparability}, we show the F1 score obtained at the last layer of two architectures: Simple \gls{fcnn} and EagerNet with the same number of layers and neurons. We use the F1 score to reduce the class imbalance bias. Results suggest that on average, EagerNet achieves similar scores to a conventional \gls{fcnn} at the last layer across all variants and datasets. It is worth mentioning, however, that EagerNet is used in this case as a conventional neural network and predictions are taken only from the last layer hence, comparing the influence of our training procedure on the typical workflow of the neural network. Using predictions from various layers by setting a confidence threshold might lead to higher accuracy if middle layers are optimized to predict certain classes.

\subsubsection{Weighted loss}
In \autoref{tab:weights} we gather results from different EagerNet networks trained using different numbers of hidden layers and weight distributions each with 128 neurons per layer. In addition, we compare the results using four metrics (Accuracy, Precision, Recall and Youden's J) and show the performance of the last layer (by setting the confidence level to the maximum). The CICIDS2017 indices %\todo{Max: Indices? Fares: it's commun to refer to acc, prec, recall etc. as indices/scores}
show that the weight distribution and the architecture itself (number of layers) have little effect on performance. In fact, since the last layer is always trained with one loss regardless of the weight distribution, only the first few layers are affected by the combined-loss learning procedure. Nonetheless, since the accuracy did not improve when increasing the number of layers, that implies that most network patterns are already separable after the few first layers which makes the classification task for CICIDS2017 manageable even for shallow neural networks. Results of UNSW-NB15 show a slightly different behavior. %\todo{Max: Sure? All the results look the same? Fares: slightly better}
The uniform weighting of losses gives best scores on average and, in addition, deeper architectures outperform the shallow ones. This implies that the network patterns in this specific dataset are complex than in the previous one therefore, requiring more non-linearities (layers) to find the optimal \emph{input:prediction} mapping function.

@@

\autoref{fig:confidence} shows the accuracy-speed balance in terms of confidence threshold. Four different networks for both datasets are shown. We experiments with two networks: 128 neurons by 3 layers and 64 neurons by 12 layers. The confidence threshold is shown on the horizontal axis, the average number of layers used over all samples and the average accuracy achieved are both shown on the vertical axis. Curves show promising and consistent results, whereby the network always uses few layers for the majority of samples and only continues to evaluate further layers if the confidence level required is too high or the sample is noisy thus it is difficult to make correct decisions in early layers. We observe moreover two phenomena:
\begin{itemize}
\item The accuracy of CICIDS2017 increases almost linearly in sections with respect to confidence and reaches its maximum when all layers are evaluated.
\item The accuracy of UNSW-NB15 is stable when the confidence level is increased from $\approx 0.62$ to until $\approx 0.98$ where it leaps noticeably once a further layer is evaluated. This suggests that some patterns were ``learned'' only on that specific layer and hence, respective samples suddenly were correctly classified.
\end{itemize}
Overall, EagerNet confirms that it is possible to trade a tiny percentage of accuracy in order to save a significant amount of resources. In section \ref{subsubsec:comparability_to_fnns}, we see that, it is feasible to achieve comparable accuracies to traditional \gls{fcnn} allowing the reduction of costs.

@@

@@

\begin{figure*}
\center
\begin{tabular}{cccc}
\subfloat[CICIDS2017 (12 layers $\times$ 64 neurons).]{\includegraphics[width=0.55\columnwidth]{figures/multiclass17.pdf}} &
\subfloat[CICIDS2017 (5 layers $\times$ 128 neurons).]{\includegraphics[width=0.35\columnwidth]{figures/multiclass17-3.pdf}} 
\subfloat[UNSW-NB15 (12 layers $\times$ 64 neurons).]{\includegraphics[width=0.68\columnwidth]{figures/multiclass15.pdf}} &
\subfloat[UNSW-NB15 (5 layers $\times$ 128 neurons).]{\includegraphics[width=0.38\columnwidth]{figures/multiclass15-3.pdf}}
\end{tabular}
\caption{Accuracy achieved on a test set per layer and attack family (sorted vertically). Some attacks have poor accuracy because only few samples are observed (refer to \autoref{fig:datasets}). Forcing the network to learn those samples might lead to over-fitting. Additionally, the effect of over-sizing the neural network immediately appears and allows to apply the EagerNet strategy.}
\label{fig:multiclass}
\end{figure*}
